---
output: html_document
---

```{r setup-01, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)

source("./functions/helper.R")
```

# Download network data {#session-download-data}

This <a href="https://www.web-of-life.es" target="_blank">web interface</a> provides a visualization of where data of a given network have been collected over the globe. 
Clicking the the dot corresponding to a network, the relative metadata can be visualized as well as a 3D representation of the network. 
The incidence matrix is also displayed and can be downloaded in different formats.  
Networks are named according to the following convention 
<!-- specifications are contained within the name of each network -->
  
* **interaction type**:  mutualistic (M) or antagonistic (A) 
* **network type**: Plant-Ant (PAO), Plant-Epiphyte (PE), Pollination (PL), Seed Dispersal (SD), Food Webs (FW), Host-Parasite (HP), Host-Parasitoid (HPD), Plant-Herbivore (PH), Anemone-Fish (AF)
* **id** labeling a specific network. 


## Download one network  

In a research projects it is often convenient to bypass the web interface and store data directly in a dataframe using a command line. This has several advantages in terms of efficiency and reproducibility of the analysis you are conducting. We will see how to do this in `R` using the function `fromJSON()` of the package  `jsonlite`, but similar commands can be run in `python` or in other languages. To this aim, we have opened few dedicated endpoints on the Application Programming Interface (API) of the  web-of-life project. 

Let us focus on the network number **002** of the **Seed Dispersal** type, named **M_SD_002**: the set fo commands to download it are
```{r}
# define the base url
base_url <- "https://www.web-of-life.es/"     
# define the full url associated with the endpoint that download networks interactions
json_url <- paste0(base_url,"get_networks.php?network_name=M_SD_002")
# import the jsonlite package (if not done at the beginning) 
library(rjson) 
# and run it 
SD_002_nw <- jsonlite::fromJSON(json_url)
```
It is always good practice to have a look at the class  of a newly defined object (dataframe) and display its first rows to understand how this data looks like, number of columns, their names, etc. 
```{r}
# check the class 
class(SD_002_nw)

# have a look at the dataframe
head(SD_002_nw)
```
The content of a dataframe can be visualized more nicely using the package `formattable`
```{r}
# import the formattable package (if not done at the beginning) 
library(formattable)
# visualize the dataframe in nicer way
formattable(head(SD_002_nw)) # visualize the dataframe in a nicer way

# formattable(SD_002_nw) # to see all the rows
```


## Download multiple networks  

If you hit the endpoint `https://www.web-of-life.es/get_networks.php` without passing any argument into the url, all the networks available in the database are returned, namely a dataframe with columns 
*network_name, species1, species2, connection_strength*: 
```{r}
json_url <- paste0(base_url,"get_networks.php") 
all_nws <- jsonlite::fromJSON(json_url)

# show results 
formattable(head(all_nws))
```
You can list and count all the distinct networks you have downloaded using the function `distinct()` of the  <a href="https://dplyr.tidyverse.org/"  target="_blank">dplyr</a> package for data manipulation: 
```{r, warning=FALSE, message=FALSE}
# import the dplyr package (if not done at the beginning) 
library(dplyr)
all_nw_names <- distinct(all_nws, network_name)

# show  results 
formattable(head(all_nws))
```

Another useful functionality provided by the <a href="https://dplyr.tidyverse.org/"  target="_blank">dplyr</a> package is the *pipe operator* `%>%` which allows passing the output of a function to another one. For instance, 
```{r, eval=FALSE}
formattable(head(all_nws))
# is equivalent to 
head(all_nws) %>% formattable()
```

Networks can be filtered by **interaction type** passing this as an argument in the url. Accepted values of the parameter `interaction_type` are:  <br>
  * Anemone-Fish  <br>
  * FoodWebs <br>
  * Host-Parasite <br> 
  * Plant-Ant <br>
  * Plant-Herbivore  <br>
  * Pollination  <br>
  * SeedDispersal <br>
  * [default ALL]  

Here the example for pollination networks: 
```{r}
json_url <- paste0(base_url,"get_networks.php?interaction_type=Pollination") 
pol_nws <- jsonlite::fromJSON(json_url)
```


Networks can also be filtered by **data type** setting the parameter `data_type` equals to  weighted or binary  [default ALL]. To download weighted, pollination networks you can run the command: 
```{r}
json_url <- paste0(base_url,"get_networks.php?interaction_type=Pollination&data_type=weighted") 
pol_weighted_nws <- jsonlite::fromJSON(json_url)
```


## Download network metadata  

You can obtain a csv file containing the **information about species** in a given network hitting the endpoint `https://www.web-of-life.es/get_species_info.php` and passing the network name into the url, namely

```{r}
# define the base url
base_url <- "https://www.web-of-life.es/"     
# download dataframe 
M_PL_073_info <- read.csv(paste0(base_url,"get_species_info.php?network_name=M_PL_073"))
# show results 
head(M_PL_073_info) # %>% formattable()
```

You can obtain a csv file with the **metadata** of all the networks hitting the endpoint 
`https://www.web-of-life.es/get_network_info.php`:      
```{r}
all_nw_info <- read.csv(paste0(base_url,"get_network_info.php"))

# see what information is provided 
colnames(all_nw_info)

# select only some specific columns 
my_nw_info <- select(all_nw_info, network_name, location, latitude, longitude)

head(my_nw_info) %>% formattable()

```
For completeness, we mention that passing the initial part of a network name into the url you can obtain the metadata for networks whose name starts, e.g., with *M_PL_072*
```{r}
my_nws_info <- read.csv(paste0(base_url,"get_network_info.php?network_name=M_PL_072")) %>%
  select(network_name, location, latitude,longitude)
# visualize data
my_nws_info %>% formattable()
```


### Example: networks in the tropical region {-}   

If you are interested in knowing which networks in <a href="https://www.web-of-life.es" target="_blank">Web of life</a> have been collected in the tropical region, you can use function `filter()` of the  <a href="https://dplyr.tidyverse.org/"  target="_blank">dplyr</a> package for data manipulation: 
```
tropical_nws_info <- my_nws_info %>% filter(.,between(latitude, -23.43647, 23.43647)) 
```


