---
output: html_document
---

```{r setup-02, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)

source("./functions/helper.R")
```
# Analyze networks with igraph

In the language of graph theory (or complex networks) species of an ecological network would be the **nodes** (**vertices**) of a graph and **links** (or **edges**) their pairwise interactions. 

The reference package for graph/network analysis is <a href="https://igraph.org/" target="_blank">igraph</a>. 
It is an open source software, optimized by professional developers and can be invoked from R, Python, Mathematica, and C/C++. It contains a lot of libraries, though you will explore only few of them throughout this course. 
The recommendation is that, whenever you need to do some operation or calculation with networks, you first look if that functionality is already provided in `igraph`. If not, you can probably build on some existing functions to develop your own function/package. 

In the following we will use the word *graph* meaning an igraph object in the context of the `igraph` package. But, obviously, <a href="https://en.wikipedia.org/wiki/Graph" target="_blank">graph</a> refers to a more general mathematical entity. 
Network data can be represented in several ways, corresponding to different data structures (classes). Below we will consider the following ones: 

| representation | data structure | 
|:----:|:---------------:|
| edge list | dataframe | 
| graph | igraph object | 
| adjacency matrix | matrix/array |
| incidence matrix | matrix/array | 



## From dataframe to igraph object 

The network data downloaded from <a href="https://www.web-of-life.es" target="_blank">Web of life</a> are stored in a representation (*network_name, species1, species2, connection_strength*). The last three columns correspond to the `edgelist` representation of a graph. Therefore, the values contained in those three columns can be converted into a graph object using the function `graph_from_data_frame()` of the  <a href="https://igraph.org/" target="_blank">igraph</a> package 
```{r}
# download the foodweb FW_001 
json_url <- paste0(base_url,"get_networks.php?network_name=FW_001")
FW_001_nw <- jsonlite::fromJSON(json_url)

# check the class 
class(FW_001_nw) 

# import the igraph package (if not done at the beginning) 
library(igraph)

# select the 3 relevant columns and create the igraph object 
FW_001_graph <- FW_001_nw %>% select(species1, species2, connection_strength) %>% 
  graph_from_data_frame(directed = TRUE)

# check the class 
class(FW_001_graph) 
```

As you can see from the last command, we have just created an *igraph object*. Note that, since in foodwebs it is relevant *who eats whom*, the representative graph is **directed**. This is specified setting the corresponding  parameter as `TRUE` in the function `graph_from_data_frame()`.  
 


## Plot a foodweb as a directed graph

Once the network is represented as an `igraph` object it  can easily  be visualized 

```{r} 
plot(FW_001_graph, vertex.size=4, 
     vertex.label=NA, 
     edge.arrow.size= 0.3, 
     layout = layout_on_sphere)  
```

Other possible layouts are: layout_nicely, layout_with_kk, layout_randomly, layout_in_circle, layout_on_sphere.  



## From igraph object to adjacency matrix 

Another way to represent an ecological network is the 
<a href="https://en.wikipedia.org/wiki/Adjacency_matrix#:~:text=The%20adjacency%20matrix%20of%20a,the%20degree%20of%20each%20vertex." target="_blank">adjacency matrix</a>, that is a square matrix whose rows and columns run over all the species name in the network and whose elements indicate the strength of interaction between species pairs (zero means no interaction). This data structure is generally less efficient from the perspective of memory storage w.r.t. an edge list, but it can be useful to facilitate visualization or to perform some calculations (e.g., evaluate the nestedness of a network, perform spectral analysis, etc.). 

Using the function `as_adjacency_matrix()` of the <a href="https://igraph.org/" target="_blank">igraph</a> package 
the adjacency matrix associated with a given network can be created.  Let us focus on the relatively small **FW_002** foodweb. Some care must be put in passing the right arguments to the function `as_adjacency_matrix()`  
because the graph is directed. The reader is addressed to the 
<a href="https://igraph.org/r/doc/as_adjacency_matrix.html" target="_blank">user manual</a> for more details. 


```{r}

# download the foodweb FW_002
json_url <- paste0(base_url,"get_networks.php?network_name=FW_002")
FW_002_nw <- jsonlite::fromJSON(json_url)

# select the 3 relevant column and pass and create the igraph object 
FW_002_graph <- FW_002_nw %>% select(species1, species2, connection_strength) %>% 
  graph_from_data_frame(directed = TRUE)

# convert the igraph object into adjacency matrix 
adj_matrix <- as_adjacency_matrix(FW_002_graph, 
                                     type = "lower", # it is directed!!!
                                     attr = "connection_strength",
                                     sparse = FALSE)  

```
If you give a quick look at the variable *adj_matrix* running the command 
```
 head(adj_matrix)
```
you will notice that it does not produce exactly what we want: 1. elements are strings, instead of numerical values, and 2. there are some empty strings to indicate that a pair of species do not interact. We can fix both problems as follows     
```{r}
# convert elements into numeric values 
class(adj_matrix) <-"numeric"

# remove NA values (previous empty strings)
adj_matrix[which(is.na(adj_matrix) == T)]<-0

# check dimensions and compare with the number of species on the Web interface 
dim(adj_matrix)
```


```{r}
# convert the adjacency matrix into a dataframe just for visualization purposes 
df <- adj_matrix %>% as.data.frame() 

df <- df[order((rownames(df))),] # to order by column names 

# visualize the adjacency matrix 
df %>% formattable()
```




## Bipartite networks {#bipartite-nw-02}

Several ecologically relevant networks are **bipartite**, that is <br>
1. interactions are not directed <br>
2. species can be split in two groups and interactions occurs only from one group to the other.   

Typical examples are *pollination* or *seed-dispersal* networks. Let us see how these features of bipartite networks affect the way in which their data has to be handled with <a href="https://igraph.org/" target="_blank">igraph</a>.  
We first  download the seed-dispersal network **M_SD_002** and convert into an igraph object: 
```{r}
# download the foodweb FW_002
json_url <- paste0(base_url,"get_networks.php?network_name=M_SD_002")
M_SD_002_nw <- jsonlite::fromJSON(json_url)

# select the 3 relevant columns and create the igraph object 
M_SD_002_graph <- M_SD_002_nw %>% select(species1, species2, connection_strength) %>% 
  graph_from_data_frame() 
```

To turn this object into a **bipartite** graph, one needs to determine which nodes (species) of the graph **M_SD_002_graph** belong to one or the other guild, e.g., plants or animals.  
Within each network we label species acting as **resources** and opposed them to the  **consumers**. Technically, this information is returned by the boolean `is.resource` when the endpoint `https://www.web-of-life.es/get_species_info.php` is hit. 
For clearness, let us store this information in the list `isResource`  

```{r}
M_SD_002_info <- read.csv(paste0(base_url,"get_species_info.php?network_name=M_SD_002"))
isResource <- M_SD_002_info$is.resource %>% as.logical() # 0/1 converted to FALSE/TRUE
```

This list can be used to split the nodes of the graph **M_SD_002_graph** into two ecologically meaningful guilds;  <a href="https://igraph.org/" target="_blank">igraph</a> uses the `type` (TRUE/FALSE) attribute of nodes to identify the two groups of species. In code,  this reads 

```{r}
# Add the "type" attribute to the vertices of the graph 
V(M_SD_002_graph)$type <- !(isResource) 
```

By convention, when representing the graph **M_SD_002_graph** as an **incidence** matrix, we want resources to be associated with rows. These species are label with `is.resource=TRUE`. Instead, `igraph` displays as rows of the **incidence** matrix nodes with the attribute `type=FALSE`. For this reason, we applied the negation operator `!()` in the previous chuck.  

One can also check that the igraph object was successfully converted into a bipartite graph running the following command  
```{r}
# check if igraph understands this object as bipartite 
is_bipartite(M_SD_002_graph)
```
It is possible to pass from the representation of a bipartite graph to that of 
<a href="https://en.wikipedia.org/wiki/Incidence_matrix#Undirected_and_directed_graphs" target="_blank">incidence_matrix</a>. 
The latter is a matrix whose rows represent one group of species (e.g. plants) and  columns the other group (e.g. animals).
This can be achieved using the function `as_incidence_matrix()` of the 
<a href="https://igraph.org/" target="_blank">igraph</a> package. 
*Mutatis mutandis* we first repeat the same steps done for the conversion from an igraph object to adjacency matrix 
```{r}
# convert the igraph object into incidence matrix 
inc_matrix <- as_incidence_matrix(
  M_SD_002_graph,
  attr = "connection_strength",
  names = TRUE,
  sparse = FALSE
)

# convert elements into numeric values 
class(inc_matrix) <-"numeric"

# remove NA values (previous empty strings)
inc_matrix[which(is.na(inc_matrix) == T)]<-0

# check dimensions and compare with the number of species on the Web interface 
dim(inc_matrix)
```

We can now visualize the elements of the incidence matrix with the function `formattable()` and order alphabetically rows and columns by relative names: 
```{r}
# convert the incidence matrix into a dataframe just for visualization purposes 
df <- inc_matrix %>% as.data.frame() 

df <- df[,order((colnames(df)))] # to order by column names 
df <- df[order((rownames(df))),] # to order by row names 

# visualize the adjacency matrix 
df %>% formattable()
```

